Measurement of total hadronic differential cross sections in the LArIAT experiment

### 2018-04-15 Chapter-2/2-LArTPCs.tex
s

\chapter{Liquid Argon Detectors at the Intensity Frontier}\label{ch:1}

@@

%Bubble-chamber experiments played a key role in probing the properties of ?-interactions. The Liquid Argon Time Projection Chambers (LArTPC) technology,  first proposed by C.Rubbia in 1977 with ICARUS project [14], is considered the modern evolution of bubble-camber concept, with the additional features of three-dimensional event reconstruction, high-resolution calorimetry, active mass coincident with detector sensitive mass and can intrinsically supply a trigger signal (self-triggering) by means of the scintillation light produced in the liquid noble gas. This technology is ideal to perform $\nu$-studies in a broad energy range, from MeV up to few GeV, with high event reconstruction efficiency, thanks to the capability of particle identifcation and detailed reconstruction of different interaction topologies. In Figure 1.4 is shown a neutrino interaction event, producing a proton, a pion and a muon, as seen in a bubble chamber and in a LArTPC.

\subsection{Time Projection Chamber}
\subsection{Ionization Detectors with Noble Liquids}
\subsection{LArTPC: Principles of Operation}
\subsection{Liquid Argon Ionization Charge Detection}
\subsubsection{Electron Life Time \& purity}
\subsubsection{Space Charge Effect}
\subsubsection{Recombination Effect}
\subsection{Liquid Argon scintillation Light Detection}
\subsubsection{LAr Scintillation Process}
\subsubsection{Wavelength Shifting of LAr Scintillation Light}
\section{The SBN Program: Neutrino Interaction and Detection}
%\subsection{SBN Goals}
%\subsection{Neutrino Interactions and Detection }
\section{DUNE: Rare Decay Searches}
The key elements for a rare decay experiment are: massive active volume, long exposure, high identification efficiency and low background. 
%The limit to proton lifetime in case of absence of signal and backgrounds is set by calculating
%$$\tau/B > M\times \epsilon\times T \times 10^{32},$$ 
%where M is the detector mass in kton, $\epsilon$ the signal detection efficiency after cuts to suppress backgrounds (dependent on the considered decay mode), T is the exposure in years, B the assumed branching fraction for the considered mode and  $10^{32}$ is a factor accounting for the number of nucleons in a kton of material \cite{Bueno2007}.
Figure \ref{fig:PDKExperimentalLImit} shows the current best experimental limits on nucleon decay lifetime over branching ratio (dots). Historically, the dominant technology used in these searches has been water Cherenkov detectors: all the best experimental limits on every decay mode are indeed set by Super-Kamiokande \cite{PhysRevD.90.072005,PhysRevLett.115.121803}.  It is particularly important to notice that the kaon energy for the proton decay mode $p \rightarrow K^+ \bar{\nu}$ is under Cherenkov threshold.  Super-Kamiokande set the limit on the lifetime for the $p \rightarrow K^+ \bar{\nu}$ mode by  relying exclusively on photons from nuclear de-excitation. For this reason, an attractive alternative approach to identifying nucleon decay is the use of a Liquid Argon Time Projection Chamber (LArTPC).

>>>>

\chapter{Liquid Argon Detectors at the Intensity Frontier}\label{ch:2}

@@

\subsection{Time Projection Chamber}
\subsection{Ionization Detectors with Noble Liquids}
\subsection{LArTPC: Principles of Operation}
\subsection{Liquid Argon Ionization Charge Detection}
\subsubsection{Electron Life Time \& purity}
\subsubsection{Space Charge Effect}
\subsubsection{Recombination Effect}
\subsection{Liquid Argon scintillation Light Detection}
\subsubsection{LAr Scintillation Process}
Liquid argon emits scintillation light at the passage of charge particle. The light yield depends on the argon purity, the electric field, the dE/dx and particle type, averaging at the tens of thousands of photons per MeV. Light emission peaks in the ultraviolet at a 128 nm, shown in comparison to Xenon and Kypton in Figure  [183].

The de-excitation of Rydberg dimers in the argon is responsible for the scintillation light.
Rydberg dimers exist in two states:  a singlet and a triplet. The time constant for the singlet radiative decay is 6 ns, resulting in a prompt component for the scintillation light. The decay of the triplet is delayed by intersystem crossing, producing a slow component with a time constant of $\sim$~1500 ns.  ``Self-trapped exciton luminescence" and  ``recombination luminescence" are the two processes responsible for the creation of the Rydberg dimers. In the first process, a charged particle excites an argon atom which becomes self-trapped in the surrounding bulk of argon,  forming a dimer; the dimer is in the singlet state 65\% of the times and in the triplet state 35\% of the times. In case of recombination luminescence, the charged particle transfers enough energy to ionize the argon. The argon form a charged argon dimer state, which quickly recombines with the thermalized free electron cloud. Excimer states are produced in the recombination, roughly half in the singlet and half in the triplet state. The light yield dependency on the electric field, on the dEdx and particle type derives from the role of free charge in the recombination luminescence process. The separation between the argon ions and the free electron cloud depend on the electric field. On one hand, a strong electric field diminishes the recombination probability, leading to a smaller light yield; on the other, it increases the free charge drifting towards the anode plane. Hence, charge and light as a function of the electric field are anti-correlated.

%The recombination rate is proportional to the local densities of both free electrons and
%positive ions, so the square of the local ionization density. Recombination is therefore
%enhanced for highly ionizing particles [102]. This results in a characteristic dE/dx
%dependence in the time profile of scintillation light [106], which can be used for pulseshape-based particle identification [186], [187].

%Although attempts to measure the absorption length of pure argon at 128 nm
%have been hindered by the presence of impurities [188], it is expected to be much
%longer than the length scale of existing or proposed LArTPCs.

The production mechanism through emission from bound excimer states implies that argon is  transparent  to its own scintillation light: the emitted photons are not energetic enough to re-excite the argon bulk.
%Scattering processes, on the other hand, do occur on length scales comparable to
%the size of existing detectors. In particular, Rayleigh scattering of 128 nm light
%from thermal density fluctuations in the liquid has a predicted scattering length of
%approximately 90 cm [190]. Scattering causes the light propagating between a source
%and optical detector to take a longer path than the direct line of sight, and affects the
%arrival times and directions of detected photons. This is particularly important for ??0
%determination in a LArPC. A direct measurement of the Rayleigh scattering length
%has been made [191] and a value 66 cm was obtained, shorter than the theoretical
%prediction.

\subsubsection{Wavelength Shifting of LAr Scintillation Light}
\subsection{Signal processing}
\section{The SBN Program: Neutrino Interaction and Detection}
%\subsection{SBN Goals}
%\subsection{Neutrino Interactions and Detection }
\section{DUNE: Rare Decay Searches}
The key elements for a rare decay experiment are: massive active volume, long exposure, high identification efficiency and low background. 
%The limit to proton lifetime in case of absence of signal and backgrounds is set by calculating
%$$\tau/B > M\times \epsilon\times T \times 10^{32},$$ 
%where M is the detector mass in kton, $\epsilon$ the signal detection efficiency after cuts to suppress backgrounds (dependent on the considered decay mode), T is the exposure in years, B the assumed branching fraction for the considered mode and  $10^{32}$ is a factor accounting for the number of nucleons in a kton of material \cite{Bueno2007}.
Figure \ref{fig:PDKExperimentalLImit} shows the current best experimental limits on nucleon decay lifetime over branching ratio (dots). Historically, the dominant technology used in these searches has been water Cherenkov detectors: all the best experimental limits on every decay mode are indeed set by Super-Kamiokande \cite{PhysRevD.90.072005,PhysRevLett.115.121803}.  It is particularly important to notice that the kaon energy for the proton decay mode $p \rightarrow K^+ \bar{\nu}$ is under Cherenkov threshold.  Super-Kamiokande set the limit on the lifetime for the $p \rightarrow K^+ \bar{\nu}$ mode by  relying exclusively on photons from nuclear de-excitation. For this reason, an attractive alternative approach to identifying nucleon decay is the use of a Liquid Argon Time Projection Chamber (LArTPC).

### 2018-04-15 Chapter-7/7-BackgroundSub.tex
s

\chapter{Background subtraction}
\section{Assessing Beamline Contamination}
What is the beamline contamination? We define beamline contamination every TPC track matched to the WC track which is not a primary pion. There are 4 different types of beamline contaminations:
\begin{itemize}
\item[]1) electrons,
\item[]2) muons,
\item[]3) secondaries from pion events,
\item[]4) matched pile up events.
\end{itemize}

So, how do we handle this contamination?
The first step is to estimate what percentage of events used in the cross section calculation is not a primary pion.

@@

\begin{figure}
\includegraphics[width=0.5\textwidth,height=\textheight,keepaspectratio]{Chapter-7/Images/Beam60A.png}
\includegraphics[width=0.5\textwidth,height=\textheight,keepaspectratio]{Chapter-7/Images/Beam100A.png}
\label{fig:BeamComposition}
\caption{Beam composition for the -60A runs (left) and -100A runs (right). The solid blue plot represents the pion content, the yellow plot represents the muon content and the grey plot represents the electron content. The plots are area normalized to the number of data events, shown in red. }
\end{figure}

Table \ref{tab:beamline} shows the beam composition per magnet setting after the mass selection according to the G4simulation, 
\begin{table}[]
\centering
\label{tab:beamline}
\begin{tabular}{|l|c|c|}
\hline
                     & I = -60 A           & I = -100 A \\ \hline
G4Pions       &   68.8 \%           &      87.4 \%        \\ \hline
G4Muons     &     4.6 \%           &        3.7 \%         \\ \hline
G4Electrons &   26.6 \%           &        8.9 \%        \\ \hline
\end{tabular}
\caption{Beamline composition per magnet settings}
\end{table}

\begin{table}[]
\centering
\label{tab:databreakdown}
\begin{tabular}{|l|c|c|c|c|c|}
\hline
                                                     & I = -60 A          & I = -100 A   & Total     &  w$_{60A}$ &  w$_{100A}$\\ \hline
Data events after Mass Cut         &     70192          &  76056       & 146248 & 0.52 & 0.48\\ \hline
Data events for Cross Section     &                         &                   &              & & \\ \hline
\end{tabular}
\caption{Data events per magnet settings}
\end{table}

We calculate the electron to pion and muon to pion ratio on the whole sample as the weighted sum of the corresponding ratio in the two current settings, 
\begin{equation}
\frac{N_e}{N_\pi}_{Data} = w_{60A}\frac{N_e}{N_\pi}_{60A}  + w_{100A}\frac{N_e}{N_\pi}_{100A},
\end{equation}
\begin{equation}
\frac{N_\mu}{N_\pi}_{Data} = w_{60A}\frac{N_\mu}{N_\pi}_{60A}  + w_{100A}\frac{N_\mu}{N_\pi}_{100A},
\end{equation}
where the weights $w_{60A}$ and $w_{100A}$ are the percentage of events in the corresponding magnet configuration passing the mass selection in data, as shown in \ref{tab:databreakdown}.

Once the beam composition is know,  we simulate the electrons, muons and pions with the DDMC and we subject the three samples to the same selection chain (WC2TPC match, shower filter, pile up filter, etc...). The percentage of electrons and muons surviving the selection chain is the  electron and muon contamination in the pion cross section sample.

\subsection{Contamination from secondaries}
The percentage of secondaries is given in the MC by the number of matched WC2TPC tracks which are not flagged as primary by Geant4.
We estimate the last type of contamination, the ``matched pile up" events, to be a negligible fraction, because of the definition of the WC2TPC match: we deem the probability of a single match with a halo particle in the absence of a beamline particle\footnote{ Events with multiple WC2TPC matches are always rejected.} extremely small.

\section{Subtraction}
Once we estimate the contaminants to primary pion ratio, the next step is subtracting their contribution from data for each type of contaminant independently. The contaminant samples are reconstructed and the corresponding interacting and incident histograms are produced. We then perform a bin by bin subtraction in the data interacting and incident histograms separately. A graphical rendering of this procedure is shown in Fig \ref{fig:backgroundSubtraction}
Once the data is background subtracted, we apply the correction laid out in the previous section.
\textcolor{blue}{How do we account for the error in the contamination subtraction? We change the electron/pion and muon/pion ratio and we see how much difference we get?}

@@

>>>>

\chapter{Background subtraction}
\section{Assessing Beamline Contamination}
Even if pions are by far the biggest beam component in negative polarity runs, the LArIAT beam is not a pure pion beam. While useful to discriminate between pions, kaons, and protons, the beamline detectors are not sensitive enough to  discriminate among the lighter particles in the beam: electrons, muons and pions fall under the same mass hypothesis. Thus, we need to assess the contamination from beamline particles other than pions in the event selections used for the pion cross section analysis and correct for its effects.

We define beamline contamination every TPC track matched to the WC track which is not a primary pion. Potentially, there are 4 different types of beamline contaminations:
\begin{itemize}
\item[]1) electrons,
\item[]2) muons,
\item[]3) secondaries from pion events,
\item[]4) matched pile up events.
\end{itemize}

So, how do we handle this contamination?
The first step is to estimate what percentage of events used in the cross section calculation is not a primary pion.  The next two sections will illustrate this estimate for the electrons, muons and secondaries from pion event.
We estimate the last type of contamination, the ``matched pile up" events, to be a negligible fraction, because of the definition of the WC2TPC match: we deem the probability of a single match with a halo particle in the absence of a beamline particle\footnote{ Events with multiple WC2TPC matches are always rejected.} negligibly small.

@@

\begin{figure}
\includegraphics[width=0.5\textwidth,height=\textheight,keepaspectratio]{Chapter-7/Images/Beam60A.png}
\includegraphics[width=0.5\textwidth,height=\textheight,keepaspectratio]{Chapter-7/Images/Beam100A.png}
\caption{Beam composition for the -60A runs (left) and -100A runs (right). The solid blue plot represents the simulated pion content, the yellow plot represents the simulated muon content and the grey plot represents the simulated electron content. The plots are area normalized to the number of data events, shown in red. }
\label{fig:BeamComposition}
\end{figure}

Table \ref{tab:beamline} shows the beam composition per magnet setting after the mass selection according to the G4Beamline simulation.
\begin{table}[]
\centering
\begin{tabular}{|l|c|c|}
\hline
                     & I = -60 A           & I = -100 A \\ \hline
G4Pions       &   68.8 \%           &      87.4 \%        \\ \hline
G4Muons     &     4.6 \%           &        3.7 \%         \\ \hline
G4Electrons &   26.6 \%           &        8.9 \%        \\ \hline
\end{tabular}
\caption{Simulated beamline composition per magnet settings}
\label{tab:beamline}
\end{table}

\begin{table}[]
\centering
\begin{tabular}{|l|c|c|c|c|c|}
\hline
                                                     & I = -60 A          & I = -100 A   & Total     &  w$_{60A}$ &  w$_{100A}$\\ \hline
N Data Events after Mass Selection     &     70192          &  76056       & 146248 & 0.48 & 0.52\\ \hline
%Data events for Cross Section          &                         &                   &              & & \\ \hline
%Estimated e in the beamline             &   18671.1         &    6769.0  &              & & \\ \hline
%Estimated $\mu$ in the beamline     &     3228.8         &     2814.1     &              & & \\ \hline
%Estimated $\pi$ in the beamline       &    48292.1        &   66472.9   &              & & \\ \hline
\end{tabular}
\caption{Number of data events which fit the pion mass hypothesis as a function of magnet settings. The last two columns represent the faction of the data in the given magnet setting.}
\label{tab:databreakdown}
\end{table}

We calculate the electron to pion, as well as the muon to pion ratio on the whole sample as the weighted sum of the corresponding ratio in the two current settings, 
\begin{equation}
\frac{N_e}{N_\pi}_{Data} = w_{60A}\frac{N_e}{N_\pi}_{60A}  + w_{100A}\frac{N_e}{N_\pi}_{100A},
\end{equation}
\begin{equation}
\frac{N_\mu}{N_\pi}_{Data} = w_{60A}\frac{N_\mu}{N_\pi}_{60A}  + w_{100A}\frac{N_\mu}{N_\pi}_{100A},
\end{equation}
where the weights $w_{60A}$ and $w_{100A}$ are the percentage of events in the corresponding magnet configuration passing the mass selection in data, as shown in table \ref{tab:databreakdown}. Figure \ref{fig:BeamComposition} shows the mometum predictions from G4Beamline overlaid with data for the 60A runs (left) and for the 100A runs (right). The predictions for electrons, muons and pions have been staggered and their sum is area normalized to data. Albeit not perfect, these plots show a reasonable agreement between the momentum shapes in data and MC. We attribute  the difference in shape to the lack of simulation of the WC efficiency in the MC which is momentum dependent and leads to enhance the number events in the center of the momentum distribution.

Once the beam composition is know,  we simulate the electrons, muons and pions with the DDMC and we subject the three samples to the same selection chain (WC2TPC match, shower filter, pile up filter). The percentage of electrons and muons surviving the selection chain weighted by the beam composition is the  electron and muon contamination in the pion cross section sample, as shown in Table \ref{tab:MCafterCutContaminants}.

\subsection{Contamination from secondaries}
Pions can travel the length of the LArIAT beamline and interact hadronically in the steel or in the non-instrumented argon upstream to the TPC front face.  One of these products can leak into the TPC and be matched with the WC track, contribuiting to the pool of events used for the cross section calculation. We call this type of particles ``secondaries" from pion events, with a terminology inspired by Geant4. 
We estimate the number of secondaries using the DDMC pion sample.  The percentage of secondaries is given by the number of matched WC2TPC tracks whose corresponding particle is not flagged as primary by Geant4 and is not a muon, to avoid double counting with the G4Beamline estimate.  The secondary to pion ratio is $X$\% in the 60A sample and $Y$\% in the 100A sample.

\section{Beamline Background Subtraction}
Once we estimate the contaminants to primary pion ratio, the next step is subtracting their collective contribution from data. To do so, we simulate the same number of electrons, muons and pions with the DDMC separately for the two magnet settings, and we apply the same selection filters on the three samples. The number of events per particle species surviving this selection is shown on table \ref{tab:MCafterCutContaminants}.

\begin{table}[]
\centering
\begin{tabular}{| l | l | l | l | l | l | l | l | }
\hline
                                                  &  $\pi^-$  60A  & $\mu^-$ 60A & $e^-$  60A & $\pi^-$ 100A  & $\mu^-$ 100A & $e^-$  100A  \\
\hline
Total Initial events                     & 334500   & 334500  & 334500  & & & \\
After Multiplicity Rejection & 331313   & 322436   &  186261& & &\\
After WC2TPC: Selection & 201458   &  285686  &  79109  & & & \\
Evts After Shower Rejection    & 191655   &  277914 &  17477    & & &\\
  \hline
     &     &   &     & & &\\
Survival rate               &57\%  &  83\%   & 5\%  &     &  &\\
     &     &   &     & & &\\
Beam Composition               &     &   &     & & &\\ 
After Selection                      &     &   &     & & &\\
                                             & 88.5\%  & 8.5\%  &  3\%   & & &\\
\hline
%         &                              &             &                      &  \\ \hline
\end{tabular}
\caption{MC selection flow per particle species.}
\label{tab:MCafterCutContaminants}
\end{table}

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{Chapter-7/Images/Staggered60A.png}
\caption{Left: staggered contributions to the interacting kinetic energy distribution for electron (grey), muons (yellow) and pion (blue) in the 60A simulation sample. Right: staggered contributions to the incident kinetic energy distribution for electron (grey), muons (yellow) and pion (blue) in the 60A simulation sample.  }
\label{fig:stag60A}
\end{figure}

We then produce the interacting and incident histograms for the events surviving the selection for both the pions and the contaminants, weighted by the estimated beam composition.

@@

We then evaluate the relative contribution of the contaminants bin by bin in the interacting and incident histograms separately. In data, we subtract this estimated relative contaminants contribution on the interacting and incident histograms  bin by bin.

We estimate the systematic uncertainty on the cross section from this subtraction procedure by varying the electron to pion and muon to pion ratio in a suitable range of values. Figure
